\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing economic inequality with tax data - Switzerland from 1945 to 2010}

### 2014-04-11 paper/subdocuments/data_and_methods.Rnw
some notes and comments

\emph{Concepts on measuring economic resources}  \\
Most studies on inequality focus on income inequality, while economic resources are ideally conceptualized with measures for income, wealth and consumption together. \\

@@

@@

@@

@@

%[ich stelle mir eine Tabelle mit verschiedenen Dimensionen vor, anhand derer eine Einordnung unterschiedlicher Datenquellen geschieht.]
% Mögliche Vergleichsdimensionen
%- Suitability of cross country comparison
%- Assessing inequality accurately
%- Implementation of common concepts of economic ressources
%- contingency for intertemporal comparision

>>>>

\emph{Concepts on measuring economic resources}  \\
Most studies on inequality focus on income inequality, while economic resources are ideally conceptualized with measures for income, wealth and consumption together. In generall it has to be stateted, that ideal measures only can be defined in context of a given reaserach question. This holds, if one thinks about the ideal measure of economic well-being, but also about the unit of analysis. Nevertheless, some more or less unified concepts exists, which can be taken as a starting point. OECD (2013) \\

@@

% an income which is adjusted for the household size is not a real income anymore. It is a measure of indivdual welfare or of the standard of living (BFS:2012)

@@

% Streichen

@@

The income of the household generally plays the major role in the living standards attained by its members, so the way earners and non-earners grouped together in households, has to be incorporated into the analysis. Especially when the the focus of interesst lays in unrevealing the role of demographic factors.

@@

% Suitability of data sources can be judeged by different criterias. crossectional or longidutional perspective. longidutional perspective allows for dynamic

%[ich stelle mir eine Tabelle mit verschiedenen Dimensionen vor, anhand derer eine Einordnung unterschiedlicher Datenquellen geschieht.]
% Mögliche Vergleichsdimensionen
%- Suitability of cross country comparison
%- Coverage of the population of interesst
%- Implementation of common concepts of economic ressources
%- contingency for intertemporal comparision

### 2014-04-11 paper/subdocuments/introduction.Rnw
some notes and comments

Economic Resources can be seen as central indicator for life chances in general and a multitude of outcomes like physical and mental health, life expectancy and crime in particular (Wilkinson and Pickett 2009). While the study of social inequality can be considered as one of the core subjects of sociology in more recent years the concern about the widening gap was addressed by global leaders (WEF 2013) and scholars alike. Empirical evidence acknowledge the supposed trend that economic inequality increased in the majority of western countries over the last decades (OECD, 2008, 2011; Gornick and Jäntti, 2013). The pattern of evolution suggests a polarization, which can be referred to as the "hollowing of the the middle class" (Alderson and Doran 2013). Households are moving towards the top and the bottom of the distribution relative to the past, which is especially problematic as the middle class can be seen as the core of western democracies or as it is stated by Stiglitz (2012,117): "our democracy is being put at peril."
\\

@@

% Lässt sich allenfalls streichen/kürzen
% Ist es die Bezeichnung offical publications sinnvoll? Oder besser govermental publications
% rf: oder "Federal publications"?

@@

@@

@@

While there are several studies about income inequality in Switzerland, the publications concerning the distribution of wealth are scarcer, albeit the distribution of wealth seems to be a relevant dimension shaping the economic well-being of individuals as well. The cross-national Data center in Luxembourg is expanding their efforts, constructing the first cross-national wealth database. But, up to the day information for Switzerland is not available. 
% ab hier wird es irgnediwe unklar
Shorrock et al. (2013) unify several datasources (including tax data) to assess the pattern of global wealth. Their databook includes figures for Switzerland. In contrast to inequality of income concerning inequality of wealth Switzerland takes a leading position \\

% the credit suisse wealth report already uses wealth tax records, but they use it only for the period from 2000-2011. They did exactly the same for Denmark and Norway, which could be interesting to compare.
% Erbschaftssteuer erwähnen und bei den Global Wealth Reports der Credit Suisse und den Studien im Forschungsantrag weitermachen.

Up to the day, Switzerland can be situated according to the actual level of income inequality in western societies as there is a huge effort to collect data which can be harmonized to comparable measures (see Luxembourg income study). However, it is unclear how the bias through non-response affects the overall measure of inequality. Likewise a long and consistent time-series allowing to identify and explore development patterns on every point of the distribution (not only the top-shares) and a time series for inequality of wealth is missing. Building on recent developments in the field of inequality research, we assess the suitability of the publicly accessible tax data to report inequality and its changes over time. First of all this includes a discussion of the accessible measures in context of a reflection about the state of the art conceptualization of economic resources as an indicator for economic well-being. Second, we summarize and apply tax data specific technics to construct suitable measures of inequality. We expand the given set by an in depth discussion of a newly applied step to handle the incomplete coverage in tax data statistics. Third, we integrate the development of Swiss inequality into the international picture.

@@

>>>>

Economic Resources can be seen as central indicator for life chances in general and a multitude of outcomes like physical and mental health, life expectancy and crime in particular (Wilkinson and Pickett 2009). While the study of social inequality can be considered as one of the core subjects of sociology in more recent years the concern about the widening gap was addressed by global leaders (WEF 2013) and scholars alike. Empirical evidence acknowledge the supposed trend that economic inequality increased in the majority of western countries over the last decades (OECD, 2008, 2011; Gornick and Jäntti, 2013;Salverda et al. 2014). Although the rise was not uniform, a common pattern seems to be identifiable, which can be referred to as the "hollowing of the the middle class" (Alderson and Doran 2013). Households are moving towards the top and the bottom of the distribution relative to the past, which is especially problematic as the middle class can be seen as the core of western democracies or as it is stated by Stiglitz (2012,117): "our democracy is being put at peril."
\\

% Allenfalls könnte man auch (Nolan und Whealan, 2014) erwähnen. "The social Impact of income inequality: Poverty, Deprivation, and Social Cohesion" Da werden die Thesen rund um die Folgen von Ungleichheit gut zusammengefasst.

@@

% Lässt sich allenfalls streichen/kürzen -> oh April2014-> streichen
% Ist es die Bezeichnung offical publications sinnvoll? Oder besser govermental publications
% rf: oder "Federal publications"?

@@

% Bei SILC einsteigen. Allenfalls eine Grafik bringen mit den unterschiedlichen Reihen.
% Daraus wird auch ersichtlich, dass keine der verfügbaren Reihen wirklich länge Zeiträume abdecken
% we should mention that Eurostat uses EU-SILC alike and show, that there is a time series for SILC-Data aswell. Additionally we should mention that the Federal Statistical Office uses the HABE statistic

% Allenfalls erwähnen, dass die Schweiz nicht Teil des GINI database ist.
http://www.gini-research.org/articles/data_2

@@

%

@@

While there are several studies about income inequality in Switzerland, the publications concerning the distribution of wealth are scarcer, albeit the distribution of wealth seems to be a relevant dimension shaping the economic well-being of individuals as well. The cross-national Data center in Luxembourg is expanding their efforts, constructing the first cross-national wealth database. But, up to the day information for Switzerland is not available.

% ab hier wird es irgnediwe unklar
Shorrock et al. (2013) unify several datasources (including tax data) to assess the pattern of global wealth. Their databook includes figures for Switzerland. In contrast to inequality of income concerning inequality of wealth Switzerland takes a leading position \\

% Albeit the awarness of an assessment of income and wealth simuntanously is rising, we focus in this paper on income inequality, which is the know as the core driver of economic inequlity
% Dann lassen sich einige Passagen streichen. Man könnte auch auf verschiedene Publikationen verweisen, welche die Bedeutung von Vermögen untersuchen -> Buch von Piketty, Müller und Schoch und dann aber ziemlich schnell darauf verweisen, dass wir uns hier auf Einkommen konzentrieren.
% Auch erste Dell Studie erwähnen -> es hat Zahlen zur Schweiz. Auch Probleme erwähnen (Fehlende Informationen zu Altersvorsorgevermögen -> hat aber niemand eine Lösung bisher.)
% Studies on inequality have to adress several thorny challenges. One has to define the centrale measure of inequality and find adquate data. Recent studies accentuate the importance of a broader concept of economic wellbeeing including measures for income, consumption and wealth alike. But each area is confronted with own measurement problems. Measure of consumptions are largely missing in tax data, albeit deductions can be understood as mandatory consumptions somehow. Measures of wealth are included in tax data, but are not connectable to the respctive income and therefore can only be analyzed separate. In danger of falling a pittraps it seems appriate to move in small steps. Therefore we focus our analysis on income, which is undoubtly a crucial indicator of economic wellbeeing.
% Also the Federal Statistical Office is increasing their effort to get sound data concering the distribution of wealth vgl. Müller und Schoch (2014).

Up to the day, Switzerland can be situated according to the actual level of income inequality in western societies as there is a huge effort to collect data which can be harmonized to comparable measures (see Luxembourg income study, EU-SILC). However, it is unclear how the bias through non-response affects the overall measure of inequality. Likewise a long and consistent time-series allowing to identify and explore development patterns on every point of the distribution (not only the top-shares) and a time series for inequality of wealth is missing. Building on recent developments in the field of inequality research, we assess the suitability of the publicly accessible tax data to report inequality and its changes over time. First of all this includes a discussion of the accessible measures in context of a reflection about the state of the art conceptualization of economic resources as an indicator for economic well-being. Second, we summarize and apply tax data specific technics to construct suitable measures of inequality. We expand the given set by an in depth discussion of a newly applied step to handle the incomplete coverage in tax data statistics. Third, we integrate the development of Swiss inequality into the international picture.

% delete passages about wealth and include passage about the comparison with survey data

@@

% Was wollen wir wissen?
% Wir müssen uns äussern zu: 
% Analyseinheiten
% Definition des Einkommens
% Berechnungsschritten

% Bei den Ergebnissen müsste es Grafiken haben zu:
% 1) Unterschiedlichen Analyseneinheiten (Mit und ohne Nuller, Normal- und Sonderfälle)
% 2) Unterschiedliche Einkommensgrössen (Reineinkommen, steuerbares Einkommen, pseudoverfügbares Einkommen (Abzüge von direkten Bundessteuern))
% 3) Vergleich mit HABE
